taxadb: A High-Performance Local Taxonomic Database Interface

### 2019-01-10 paper/manuscript.Rmd
wip

```{r}
bbs_ids <- 
bbs %>% select(species) %>%
  mutate(gbif = get_ids(species, "gbif", "prefix"),
          col = get_ids(species, "col", "prefix"),
         itis = get_ids(species, "itis", "prefix"),
         ncbi = get_ids(species, "ncbi", "prefix"),
     wikidata = get_ids(species, "wd", "prefix"),
          ott = get_ids(species, "ott","prefix") 
  )

@@

`get_ids` by default looks only for exact matches.  We can get more matches by cleaning the names first.  `clean_names()`

```{r}
clean_names(bbs$species, 
            )
```

```{r}
ids <- get_ids(clean_names(bbs$species), "ott", "prefix")
unmatched <- bbs$species[is.na(ids)]
clean_names(unmatched)
```

```{r}
unmatched <- bbs$species[is.na(bbs_ids$ott)]
length(unmatched)

drop_sp.(unmatched) %>%
  binomial_names()

clean_names(unmatched)
```

```{r}
bbs_ids %>% 
  select(-species) %>% 
  purrr::transpose() %>%
  purrr::map_dbl(function(x) sum(!is.na(x)) < 1) %>% sum()
```

```{r}
ott_ids <- ids(species_list, "ott")
```

```{r}
## Consider: Working from a focal authority:
## select all rows where focal authority did not match but another authority did
## for each matched authority, access species name associated with id, and requery

## OR consider: access accepted species name associated with any authority, requery that in original authority?
##

## OR consider? generate a combined taxonid table for all authorities
```

ITIS resolves the most names, but no authority resolves all the names provided.

```{r}
unmatched <- species_list[is.na(bbs_ids$itis)]

first_two <- function(x){ tmp <- strsplit(x, " ")[[1]]; paste(tmp[[1]], tmp[[2]]) }
drop_sp <- function(x) gsub("(\\w+)\\s+sp\\.", "\\1", x)

matched <- unmatched %>% 
  map_chr(first_two) %>% 
  drop_sp() %>%
  get_ids("itis", "prefix")

matched %>% is.na() %>% sum()

still_umatched <- unmatched[is.na(matched)]

```{r}
first_two_fuzzy <- function(x){ 
  tmp <- strsplit(x, " ")[[1]]
  paste0(tmp[[1]], " ", tmp[[2]], "%") 
}

only_genus <- function(x){ 
  purrr::map_chr(x, function(x) strsplit(x, " ")[[1]][[1]])
}
name_to_regex <- function(x){ 
  x <- gsub("(\\w+)\\s+sp\\..*", "\\1.*", x)
  # x <- gsub("(\\w+)\\s+(\\w+).*", "\\1 \\2.*", x)
  x <- gsub("(\\w+)\\s+(\\w{1,3}).*", "\\1 \\2.*", x)
  x
}
name_to_regex(still_umatched)

## subset by genus first
tmp <- 
  right_join(taxa_tbl("itis", "hierarchy"), 
           data_frame(genus = only_genus(still_umatched)),
           copy = TRUE) %>% select(id) %>% 
  left_join(taxa_tbl("itis", "taxonid"), by="id") %>%
  distinct() %>% 
  collect()

df <- data_frame(regex_name = name_to_regex(still_umatched))
i <- vapply(df$regex_name, 
            function(x) which(grepl(x, tmp$name)),
            integer(1))
tmp[i,] %>%  select(id, name) %>% distinct()

tmp[which(grepl("Rallus.*", tmp$name)), ] %>%  select(id, name) %>% distinct()

```{r}
matched <- tmp %>%   
  fuzzyjoin::regex_inner_join(
    data.frame(regex_name = name_to_regex(still_umatched)),
    by = c(name = "regex_name")
  )
```

```{r}
matches <- 
  right_join(taxa_tbl("itis", "hierarchy"), 
           data_frame(genus = only_genus(species_list)),
           copy = TRUE) %>% 
  distinct() %>% 
  collect() 
# uses smaller table filtered by genus
library(fuzzyjoin)
left_join(data_frame(species = species_list), matches)

stringdist_left_join(data_frame(species = species_list), matches)

  right_join(data_frame(species = first_two(unmatched))) %>% 
  select(id, species) %>% 
  distinct()

taxa_tbl("itis", "taxonid") %>%
  filter(name %in% c("Rallus%", "Vermivora%"), rank == "species")

taxa_tbl("itis", "hierarchy") %>% filter(genus == "Rallus") %>% select(genus, species)
```

```{r}
itis <- ids(species_list, "itis") 
col <- accepted_name(bbs_ids$itis, "itis") %>% pull(name) %>% get_ids("col")
sum(!is.na(col))
```

```{r}
unmatched <- bbs_ids %>%  
  select(-species) %>%
  purrr::map_dfc(function(x) !is.na(x)) %>% 
  rowSums() < 1

sum(!unmatched)
```

>>>>

```{r}
bbs_ids <- 
bbs %>% select(species) %>%
  mutate(gbif = get_ids(species, "gbif", "prefix"),
          col = get_ids(species, "col", "prefix"),
         itis = get_ids(species, "itis", "prefix"),
         ncbi = get_ids(species, "ncbi", "prefix"),
     wikidata = get_ids(species, "wd", "prefix"),
  #       iucn = get_ids(species, "iucn", "prefix"),
          ott = get_ids(species, "ott","prefix") 
  )

@@

`get_ids` by default looks only for exact matches.  We can get more matches by cleaning the names first.  The `clean_names()` function provides several simple string manipulations to tidy names to improve the probability of a match: missing species (specific eptithets) such as `Accipiter sp.` drop the `sp.`, allowing matches against Genus names, and intraspecific epithets such as `Colaptes auratus cafer` are dropped to binomial names, `Colaptes auratus`.  These transformations may not be appropriate for certain use cases, and should be used with care, and only once as many untransformed names as possible have already been matched.

In this example, applying this transformation on the unmatched names we can bring matches against OTT identifiers up from 665 to 740, missing only ten names:

```{r}
bbs_ott <- bbs$species %>% get_ids("ott", "prefix")
```

```{r}
cleaned <- 
  bbs$species[is.na(bbs_ott)] %>% 
  clean_names() %>% 
  get_ids("ott", "prefix")

## use ID from cleaned name only if no exact match.
bbs_ott[is.na(bbs_ott)] <- cleaned
```

We are still missing 10 species:

```{r}
still_missing <- bbs$species[is.na(bbs_ott)]
still_missing
```

Three of these are recognizably common names, which we can resolve against the common names table:

```{r}
clean_names(still_missing) %>% ids("gbif")
```

(GBIF lists *Antigone canadensis* as a synonym for *Grus canadensis*; this appears backwards as the Sandhill Crane was moved out of *Grus* after molecular analyses showed it made that taxon polyphyletic; https://en.wikipedia.org/wiki/Sandhill_crane).
